A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
changed sentence
final version

In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithm compared everything, 
and that is where we gain.

@@

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
%
We define a class of valid path decompositions which characterizes decompositions produced by our algorithm.
%
%We define a class of valid path decompositions.  This include any decomposition of rooted trees,
%and more generally is a large superset of all decomposition produced by our 
%algorithm.
%\footnote{Our construction likely works for any decomposition, however, 
%to keep the presentation simple we restrict to a subset relevant for our algorithm.}
%
%Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
%on some input.
%
%
%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
%specific path decomposition $P(T)$%
\ignore{\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

\begin{lemma}
\label{lem:pathDecomp}
Fix a path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
\end{lemma}

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that the size of $\bF_P$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

\Thm{main-lb} now follows immediately from an entropy argument, analogous to \Thm{joinLB}.
%
%
%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
%
%\begin{theorem}
%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
%Any height function decision tree that correctly computes
%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
%\end{theorem}

>>>>

In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithms compared everything, 
and that is where we gain.

@@

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
%
%We define a class of valid path decompositions.  This include any decomposition of rooted trees,
%and more generally is a large superset of all decomposition produced by our 
%algorithm.
%\footnote{Our construction likely works for any decomposition, however, 
%to keep the presentation simple we restrict to a subset relevant for our algorithm.}
%
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.
%
%
%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
%specific path decomposition $P(T)$%
\ignore{\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

\begin{lemma}
\label{lem:pathDecomp}
Fix any path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
\end{lemma}

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $P$ be any rain consistent path decomposition of some contour tree. 
There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that the size of $\bF_P$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

We now restate \Thm{main-lb}, which follows immediately from an entropy argument, analogous to \Thm{joinLB}.

\begin{theorem} 
Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
\end{theorem}

%
%
%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
%
%\begin{theorem}
%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
%Any height function decision tree that correctly computes
%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
%\end{theorem}
